Trends in the contribution of major causes of death to the black-white life expectancy gap by US state

### 2018-05-10 Manuscript/Manuscript_COD_Trends_HealthandPlace.Rmd
updated up to S&W!

```{r save-figs, echo=F, eval=F}
ggsave(filename = "../Plots/figure1a.jpeg", plot = le.gap.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure1b.jpeg", plot = le.gap.women, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure2a.jpeg", plot = cvd.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure2b.jpeg", plot = cvd.women, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure3a.jpeg", plot = can.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure3b.jpeg", plot = can.women, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure4a.jpeg", plot = ncd.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure4b.jpeg", plot = ncd.women, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure5a.jpeg", plot = inj.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure5b.jpeg", plot = inj.women, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure6a.jpeg", plot = com.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure6b.jpeg", plot = com.women, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure7a.jpeg", plot = aoc.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure7b.jpeg", plot = aoc.women, device = "jpeg", width = 12, height = 10, units = "in")
```

@@

**Summary of findings**

We examined state-level trends in six major cause-of-death contributions to the black-white life expectancy gap over a 45-year period. While we found that many states conformed to a general pattern, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation. Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the Midwest, which had been the receiving point of blacks from the South during “the Great Migration” for improved economic opportunities between 1916 and 1970.^14^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi. On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^15^ On the other hand, because greater racial equality can be achieved by either falling black mortality rates or rising white rates, the reduced life expectancy gap cannot be celebrated without a careful examination of the two individual trends.^16^

Consistent with what has been reported at the national level,^17^ we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time. Injuries continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South. This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic.^18^ Communicable disease continues to contribute to the gap in most states, especially in DC, New Jersey, and Florida. In our previous work, approximately half of the contribution of communicable disease to the gap was found to be due to HIV in 2008 nationally,^17^ suggesting that these states may increase equality through interventions targeted at reducing mortality from HIV. Cancer is also responsible for large contributions to the life expectancy gap in some states (Illinnois, Michigan, and Wisconsin) but not in others (New York, Massachusetts). Overall, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women. However, both states began with only a minimal contribution of CVD and cancer to the mortality gap. The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contributions of more than two years. Thus, the reason for New York’s and Massachusetts’ low CVD contributions does not appear to be the effect of recent changes, but highlights a long-standing pattern of greater equality and points to the importance of viewing the most recent estimates in light of historical patterns.

@@

We used data over 45 years, across two racial groups, and for all ages. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause of death were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time. Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.^19,20^ The Arriaga decomposition method accounts for the fact that deaths occurring in one cause stratum are reflected in the absence of that death in another stratum. In contrast to some recent work focused on county-level inequalities,^21,15^ states have some substantial advantages as analysis units; They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities with divergent social environments that are reflected in varying patterns of mortality by race.

Our analysis has three main limitations. Firstly, we did not account for Hispanic ethnicity, since ethnicity was not measured until 1978, and even then it was poorly measured on the death certificate in many states for many years. Our previous sensitivity analysis including Hispanics did not yield a substantial difference in the results for the overall life expectancy gap.^8^ Had Hispanics been excluded from this study, the life expectancy estimates for both blacks and whites are expected to be lower, since Hispanics are known to have lower mortality than blacks and whites.^22^ Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states. US immigrants have been shown to have life expectancies at birth that were approximately 5 years longer in 2010.^23^ Thus, the black-white gap in a state may decrease as more immigrants identify as black. Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death. Migration is particularly important for the District of Columbia, because of its small geographical bounds. Affluent black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC. Both changes would serve to increase the racial gap in DC.^24^ Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes. For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time. Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.^25,26^

**Implications for Policy & Future Work**

The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time. States pursue varying approaches to an array of relevant programs and laws including tobacco control,^27^ Medicare coverage,^28^ hand-gun violence,^29^ and nutrition.^30^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks.^31^ The analyses reported here are entirely descriptive, but could be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others. Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations.^32^

>>>>

```{r save-figs-separately, echo=F, eval=F}
ggsave(filename = "../Plots/figure1a.jpeg", plot = le.gap.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure1b.jpeg", plot = le.gap.women, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure2a.jpeg", plot = cvd.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure2b.jpeg", plot = cvd.women, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure3a.jpeg", plot = can.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure3b.jpeg", plot = can.women, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure4a.jpeg", plot = ncd.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure4b.jpeg", plot = ncd.women, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure5a.jpeg", plot = inj.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure5b.jpeg", plot = inj.women, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure6a.jpeg", plot = com.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure6b.jpeg", plot = com.women, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure7a.jpeg", plot = aoc.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure7b.jpeg", plot = aoc.women, device = "jpeg", width = 12, height = 10, units = "in")
```

@@

**Key findings**

We examined state-level trends in six major cause-of-death contributions to the black-white life expectancy gap over a period of forty-five years. While we found that many states conformed to a general pattern, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation.

By 2013, the northeastern states of Massachusetts, Rhode Island, New York, and Connecticut had eliminated or nearly eliminated their black-white differences in life expectancy for both men and women. Each of these states had smaller contributions of each cause of death to the gap compared to the overall average, however, their contributions for CVD were most markedly lower. Notably, the contribution of CVD to the gap in these states has been lower than the overall average since 1969, reflecting that their low CVD contributions are not a consequence of recent changes, but rather a long-standing pattern of greater racial equality in CVD mortality. This points to the importance of viewing the most recent estimates in light of historical patterns. Kentucky, Nevada, New Mexico, and Washington state also performed better than the nation overall, with almost all causes of death contributing less than the overall average for both men and women.

The rustbelt states of Illinois, Wisconsin, and Michigan, alongside DC had the highest black-white gaps for both men and women. These states all had much higher contributions of CVD to the black-white life expectancy gap. Injuries also made large contributions to black-white gaps in these states and DC, but for men only.  Cancer and All other causes also played a role in Wisconsin and especially in DC, for which the black-white gap in 2013 was far higher than any other state.

For women, several southern states made significant progress towards narrowing the black-white life expectancy gap. For example, Kentucky, Georgia, Alabama, and West Virginia had the smallest black-white gaps among southern states in 2013 for women, all less than two years. While injuries were not a large contributor to the black-white gap for women, its underlying race-specific mortality trends are concerning. Beginning in the mid ‘90s, trends in injury mortality increased for white women and decreased for black women, implying that a portion of the progress towards narrowing the black-white difference in several states is due to increasing injury mortality among white women. This likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic (Rudd et al., 2016).

The racial gap in life expectancy is now smaller in many states in the South Atlantic and the East South Central census divisions than in the Midwest (e.g., less than 2.5 years in South Carolina, Delaware, Florida, Mississippi, North Carolina, Maryland, Tennessee, West Virginia, Alabama, Georgia, Kentucky, vs. greater than four years in Michigan, Illinois, Kanses, Iowa, Nebraska, and Wisconsin for women). This regional comparison is of particular interest because of the Great Migration between 1916 and 1970, during which a substantial number of blacks moved from the rural south to the urban north (Lemann et al., 1991). Generally speaking, migration to areas with increased economic prospects has been linked to raising health in the recipient locations and decreasing health in the origin locations because migrants are generally healthier than non-migrants (Normal et al., 2005). However, an econometric analysis has linked the Great Migration to increased mortality among migrants (Black et al., 2015). Thus, it may be that the heightened inequalities in rustbelt states partially reflects elevated black mortality associated with migration, though the Great Migration cohorts are less reflected in the later time period.

Communicable disease continues to contribute to the gap in most states, but especially in DC, New Jersey, and Florida. In our previous work, approximately half of the contribution of communicable disease to the gap was found to be due to HIV in 2008 overall, (Harper et al., 2012) suggesting that these states may increase equality through interventions targeted at reducing mortality from HIV. Cancer is also responsible for large contributions to the life expectancy gap in some states (DC, Wisconsin, Illinois) but not in others (New York, Massachusetts, Rhode Island), with most states roughly following the overall pattern.

@@

We used data over 45 years, across two racial groups, and for all ages. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause of death were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time. Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.24,25 The Arriaga decomposition method accounts for the fact that deaths occurring in one cause stratum are reflected in the absence of that death in another stratum. In contrast to some recent work focused on county-level inequalities,26,27 states have some substantial advantages as analysis units. They have large populations to generate stable estimates, and tend to be more demographically stable over time.

Our analysis has three main limitations. Firstly, we did not account for Hispanic ethnicity, since ethnicity was not measured until 1978, and even then it was poorly measured on the death certificate in many states for many years. In order to investigate the potential impact of Hispanic ethnicity, we inspected mortality rates in five states with large Hispanic populations (California, New York, Florida, Texas, and Illinois). Excluding Hispanics tended to shift the mortality rates up for both blacks and whites (Supplementary Appendix 3), since Hispanics overall tend to have lower mortality rates than non-Hispanics.28 New York was the exception: excluding Hispanics shifted mortality upwards for blacks but not for whites. This is difficult to interpret without further study, but may reflect heterogeneity in health among Hispanic subpopulations that have distinct migration patterns across US states.29,30,31 The other exception we noted was with injuries: injury mortality rates for all whites vs. non-Hispanic whites diverged since 1999, and may reflect the lower overdose mortality among Hispanics.20 Our previous sensitivity analysis including Hispanics did not yield a substantial difference in the results for the overall life expectancy gap.4 Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states. US immigrants have been shown to have life expectancies at birth that were approximately 5 years longer in 2010.32 Thus, the black-white gap in a state may decrease as more immigrants identify as black. Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death. Migration is particularly important for the District of Columbia, because of its small geographical bounds. Affluent black residents have moved to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC. Both changes would serve to increase the racial gap in DC.33 Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes. For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time.34 Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.35,36

**Implications for Policy & Future Work**
Given the scope of this estimation exercise, we did not attempt to make causal inferences between the changes in mortality and black-white inequality to changes that occurred in policy, migration, health care access, or the multitude of other characteristics that vary on the state level. Making connections between changes in state policy and changes in health outcome is an important next step in this research. States have pursued varying approaches to an array of relevant programs and laws including tobacco control,37 Medicare coverage,38 hand-gun violence,39 and nutrition40, to name a few examples. These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks.41 The analyses reported here are entirely descriptive, but could be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others. Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations.42
